### 2019-06-26 2-Chapters/1-Chapter/chapter1.tex
Lots of things to do, oh my god.

\TODOL{Mardi soir : je dois encore tout relire et améliorer un peu !}

@@

% \TODOL{Est-ce que je dois détailler plus ? Quel financement j'ai eu (CDSN), quels autres financements m'ont aidés ?}

@@

% % Main problem
% \textbf{Main problems.}
%
At the root of the problems that motivate this thesis are the questions of global warming and the increase of the world population.
% https://en.wikipedia.org/wiki/History_of_radio
During the last 150 years, mankind has developed a wide range of different communication technologies, and since the late $1890$s wireless communications between man-made manufactured devices have been made possible, and more and more frequent in our lives.
With the avent of Internet of Things networks (IoT), billions of autonomous low-power devices are expected to be deployed worldwide, for a large range of different applications.
It is now a world-wide consensus that with the current trend of population increase and with the energy crisis, any newly deployed technology should be both cheap and energy efficient,
as well as adapted to serve a large number of people and devices.

That is why we are interested in this thesis about the possible applications of embedding a certain kind of Machine Learning algorithms (Multi-Armed Bandit algorithms),
directly into the future IoT devices, in order to improve their battery life and reduce the energy cost of IoT networks, by optimizing their wireless communications and learning to be self-organized, thanks to embedded decentralized decision making.

@@

This third kind of wireless systems are characterized by their decentralized nature: due to the really limited information that the base station can send to the devices, it is no longer possible to optimize the efficiency of the network from the centralized point-of-view of the base station.
In the present and future IoT networks, many devices of heterogeneous natures are using the same antenna for different applications.
A common hypothesis is that such IoT devices have a strong constraint on their power consumption, as most of them will be deployed without a direct power access and run on a tiny battery, which lifespan should be maximized, as typically commercial IoT companies sell a lifespan more than 10 years \cite{Centenaro16}.
Another common constraint for IoT devices is their low duty cycles, as most applications target one or a few messages to send every day, in strike opposition to the high data-rate pursued for centralized systems such as 4G/5G and WiFi.
%
Many different standards for IoT networks have been proposed,
and they consist in a specification for both the \emph{PHY}sical
and the \emph{M}edium \emph{A}ccess \emph{C}ontrol (\emph{MAC}) layer.
To quote some examples of standards, ZigBee, Z-Wave or Bluetooth are targeting for short-range communications, while LoRaWAN, SIGFOX, Ingenu or Weightless are designed for long-range communications.
We refer to the survey \cite{Centenaro16} for more details, and references in \cite{Azari18} or our recent article \cite{MoyBesson2019}.

@@

Cellular network bands are overloaded in most parts of the world, but other frequency bands (such as military, amateur radio and paging frequencies) are insufficiently utilized.
Independent studies performed in some countries confirmed that observation, and concluded that spectrum usage highly depends on time and place.
Moreover, the fixed spectrum allocation prevents rarely used frequencies, like those assigned to specific services, from being used, even when any unlicensed users would not cause noticeable interference to the assigned service.
Therefore, in the last $15$ years, regulatory bodies in the world have been considering whether to allow unlicensed users in licensed bands, if they would not cause any interference to the (paying) licensed users.
These initiatives have initiated the research on \textbf{Cognitive Radio},
and especially on a field that we detail below, \textbf{Dynamic Spectrum Access (DSA)}.
%
%
\begin{figure}[h!]
    \centering
    \includegraphics[width=0.65\textwidth,angle=90]{United_States_Frequency_Allocations_Chart_2016_The_Radio_Spectrum.pdf}
    \caption[A chart representing the allocation of radio spectrum in the USA in $2016$]{A chart representing the allocation of radio spectrum in the USA in $2016$. Only a small share of the spectrum is not yet allocated, the top-left corner in white (it is in logarithmic scale from \SI{0}{\kilo\hertz} to \SI{300}{\giga\hertz}). \copyright{} Wikimedia.}
    % https://upload.wikimedia.org/wikipedia/commons/c/c7/United_States_Frequency_Allocations_Chart_2016_-_The_Radio_Spectrum.pdf?uselang=fr
    \label{fig:1:United_States_Frequency_Allocations_Chart_2016_The_Radio_Spectrum}
\end{figure}

@@

\paragraph{From Machine Learning to Multi-Armed Bandits.}
%
We can narrow down the first field of study of this thesis, steps by steps.
Starting from the general concept of Machine Learning (ML), we focused our interest to statistical learning, then sequential learning, then Reinforcement Learning (RL), then RL with limited feedback and finally to Multi-Armed Bandits (MAB) learning.
As illustrated in Figure~\ref{fig:1:MainTypesOfMachineLearning}, there are other types of ML, that we do not focus on in this thesis.
The reference book on RL is \cite{SuttonBarto2018}, so let us quote the definition of RL given by R. Sutton and A. Barto:
\emph{``Reinforcement learning is learning what to do - how to map situations to actions - so
as to maximize a numerical reward signal. The learner is not told which actions to
take, but instead must discover which actions yield the most reward by trying them''}.
% - apprentissage statistique (machine learning) > séquentiel > par renforcement > information partielle (bandits)

@@

We illustrate this idea of a \emph{learning cycle} alternating between actions and feedback,
in the following Figure~\ref{fig:1:ReinforcementLearningCycle}.
A player (or learner) interacts with its environment by taking an action $A(t)$ (\eg, a choice in a finite set, $A(t)\in\{1,\dots,K\}$, or a vector $A(t)\in\R^d$), and then observing a reward $r(t)$, which is a certain measure of success from the environment (\eg, $r(t)\in\{0,1\}$ for binary failure/success, or $r(t)\in\R$).
The goal of the player is to maximize its reward, by trials and errors (\ie, actions and rewards).
Many real-world problems can be framed as Reinforcement Learning problems, as illustrated by the survey \cite{bouneffouf2019survey} and Section~\ref{sec:2:applicationsofStochasticMAB}, for instance learning to walk or to drive, learning to play a board or computer game, or discovering which treatment is efficient in healing a certain disease (clinical trial), etc.

@@

% Il faut expliquer ce que c'est que les bandits, rapidement, et leur application à l'OSA
More precisely, in this thesis we focus on Reinforcement Learning models, in particular the learner does not have access to the entire reaction of the environment after taking its action.
In other words, we focus on RL with limited feedback, when the player only sees the reward given by its action at every round, and not the reward that would have been given if she chose any other action.
This kind of limited feedback is called \emph{bandit information}, and we discuss the history and the concept of \emph{Multi-Armed Bandits} (MAB) in details in the next Chapter~\ref{chapter:2}.
Clinical trials and treatment discovery were historically the first applications of MAB since $1930$s, with the early work of W. Thompson \cite{Thompson33}.
%
MAB are a simple yet powerful example of the well-known exploration/exploitation dilemma:
when facing a set of $K$ actions whose effects on the environment are unknown, a learner must balance between
exploration of the unknown actions, in order to collect more information about them,
and exploitation of the best action, according to its current knowledge.
%
The MAB problem have been studied in both the machine learning and the statistics community, since the $1950$s with pioneers like H. Robbins \cite{Robbins52}, and more recently it is an active field of research, since the late $1990$s \cite{Anantharam87a,Anantharam87b,auer1995gambling,Agrawal95}.
Research on MAB has produced a vast literature in the $2000$s \cite{Auer02,Auer02NonStochastic,Audibert2009minimax} and continues to be a topic of high interest, as illustrated by the surveys and books \cite{Bubeck12,LattimoreBanditAlgorithmsBook,Slivkins2019}, and the wide range of applications of MAB in the recent years.

@@

% \subsection{Focusing on IoT networks.}
\subsection{Our model of IoT network and decentralized MAB}

@@

On the first hand, in the OSA setting, we failed to obtain positive result for the Selfish policy, as we proved that in some limited settings (\eg, $K=3$ and $M=2$), \Selfish-\UCB{} can suffer from linear regret with a small probability, and thus suffers from linear mean regret.
This result was later confirmed and analyzed by \cite{BoursierPerchet18}.
%
On the other hand, we were able to propose new algorithms for this multi-player bandit problem with sensing information, and we analyzed our proposal \MCTopM-\klUCB, to show that it achieves a finite-time logarithmic regret upper bound, improving over previous state-of-the-art results.
Our algorithm also achieves a logarithmic number of collisions and arm switches, and allows a fixed group of $M$ devices to efficiently learn to use the $M$ best channels orthogonally for almost all their up-link communications.
%
This strong theoretical result heavily depends on the presence of sensing feedback, and as such our results are not (yet) applicable to the IoT model without sensing.

@@

On the first hand, if the activation pattern of all devices could be fixed, for instance based on a centralized affectation of the devices to different time slots, then the model of multi-player bandit from Chapter~\ref{chapter:5} can be used to let all groups of $M \leq K$ devices learn an optimal orthogonal affectation in the $K$ channels (\eg, groups can be the set of devices that all emit at the same time).
%
However, even if we continue to assume a synchronized time in all the rest of the thesis,
it is hard to argue that this hypothesis of a centralized time schedule for the end-devices can be realistic, because we study the case of decentralized learning for OSA and IoT precisely in order to avoid any central control of the devices by the base station, as we explained above.

@@

@@

\begin{itemize}
\item
    \emph{Decentralized Spectrum Learning for IoT Wireless Networks Collision Mitigation},
    by Christophe Moy \& \textbf{Lilian Besson}.
    1st International ISIoT workshop,
    % \footnote{~See \href{https://sites.google.com/view/ISIoT2019}{\texttt{sites.google.com/view/ISIoT2019}}},
    at \emph{Conference on Distributed Computing in Sensor Systems},
    % \footnote{~IEEE DCOSS 2019, see \href{http://2019.dcoss.org}{\texttt{2019.dcoss.org}}},
    Santorini, Greece, May 2019.
    % \href{https://HAL.Inria.fr/hal-02144465}{\texttt{HAL.Inria.fr/hal-02144465}}.
    See Section~\ref{sec:4:gnuradio}.
    \cite{MoyBesson2019}

@@

\item
    \emph{Upper-Confidence Bound for Channel Selection in LPWA Networks with Retransmissions},
    by Rémi Bonnefoi, \textbf{Lilian Besson}, J. Manco-Vasquez \& Christophe Moy.
    1st International MOTIoN workshop,
    % \footnote{~MOTIoN 2019, see \href{https://sites.google.com/view/wcncworkshop-motion2019}{\texttt{sites.google.com/view/wcncworkshop-motion2019}}},
    at \emph{IEEE WCNC}, Marrakech, Morocco, April 2019.
    % \href{https://HAL.Inria.fr/hal-02049824}{\texttt{HAL.Inria.fr/hal-02049824}}.
    See Section~\ref{sec:4:retransmissions}.
    \cite{Bonnefoi2019WCNC}

\item
    \emph{GNU Radio Implementation of MALIN: ``Multi-Armed bandits Learning for Internet-of-things Networks''},
    by \textbf{Lilian Besson}, Rémi Bonnefoi \& Christophe Moy.
    \emph{Wireless Communication and Networks Conference},
    % \footnote{~IEEE WCNC 2019, see \href{http://wcnc2019.ieee-wcnc.org}{\texttt{wcnc2019.ieee-wcnc.org}}},
    Marrakech, Morocco, April 2019,
    % \href{https://HAL.Inria.fr/hal-02006825}{\texttt{HAL.Inria.fr/hal-02006825}}.
    See Section~\ref{sec:4:gnuradio}.
    \cite{Besson2019WCNC}

\item
    \emph{Multi-Player Bandits Revisited},
    by \textbf{Lilian Besson} \& Emilie Kaufmann.
    \emph{Algorithmic Learning Theory},
    % \footnote{~ALT 2018, see \href{http://www.cs.cornell.edu/conferences/alt2018}{\texttt{www.cs.cornell.edu/conferences/alt2018}}},
    Lanzarote, Spain, April 2018,
    % \href{https://HAL.Inria.fr/hal-01629733}{\texttt{HAL.Inria.fr/hal-01629733}}.
    See Chapter~\ref{chapter:5}.
    \cite{Besson2018ALT}

\item
    \emph{Aggregation of Multi-Armed Bandits learning algorithms for Opportunistic Spectrum Access},
    by \textbf{Lilian Besson}, Emilie Kaufmann \& Christophe Moy.
    \emph{Wireless Communication and Networks Conference},
    % \footnote{~IEEE WCNC 2018, see \href{http://wcnc2018.ieee-wcnc.org}{\texttt{wcnc2018.ieee-wcnc.org}}},
    Barcelona, Spain, April 2018,
    % \href{https://HAL.Inria.fr/hal-01705292}{\texttt{HAL.Inria.fr/hal-01705292}}.
    See Chapter~\ref{chapter:25}.
    \cite{Besson2018WCNC}

@@

\begin{center}
    \textbf{Copyright 2016-2019, \copyright ~Lilian~Besson.}
\end{center}

>>>>

@@

@@

% % Main problem
% \textbf{Main problems.}
%
The root of the problems motivating this thesis are the questions of global warming and the increase of the world population.
% https://en.wikipedia.org/wiki/History_of_radio
During the last $150$ years, mankind has developed many different communication technologies, and since the late $1890$s, wireless communications between man-made manufactured devices have been made possible, and more and more frequent in our lives.
With the avent of Internet of Things networks (IoT), billions of autonomous low-power devices are expected to be deployed worldwide, for a large range of different applications.
It is now a world-wide consensus that with the current trend of population increase and with the energy crisis, any newly deployed technology should be both cheap and energy efficient,
as well as adapted to serve a large number of people and devices.
%
Such IoT technologies should be able to adapt automatically to different environments and application contexts, and be as efficient as possible.
Instead of devoting a lot of time, energy and money on Research and Development to set up efficient radio access schemes, for all possible cases, a cheap and promising direction is to use Machine Learning to (try to) attain the same performance gain automatically.

That is why we are interested in this thesis about the possible applications of embedding a certain kind of Machine Learning algorithms (Multi-Armed Bandit algorithms),
directly into the future IoT devices, in order to improve their battery life and reduce the energy cost of IoT networks, by optimizing their wireless communications and by learning to be self-organized, thanks to embedded decentralized decision making.

@@

This third kind of wireless systems are characterized by their decentralized nature: due to the really limited information that the base station can send to the devices, it is no longer possible to optimize the efficiency of the network from the centralized point-of-view of the base station.
In the present and future IoT networks, many devices of heterogeneous natures are using the same antenna for different applications.
A common hypothesis is that such IoT devices have a strong constraint on their power consumption, as most of them will be deployed without a direct power access and run on a tiny battery, which lifespan should be maximized, as typically commercial IoT companies sell a lifespan more than $10$ years \cite{Centenaro16}.
Another common constraint for IoT devices is their low duty cycles, as most applications target one or a few messages to send every day, in strike opposition to the high data-rate pursued for centralized systems such as 4G/5G and WiFi.
%
Many different standards for IoT networks have been proposed,
and they consist in a specification for both the \emph{PHY}sical
and the \emph{M}edium \emph{A}ccess \emph{C}ontrol (\emph{MAC}) layer.
To quote some examples of standards, ZigBee, Z-Wave or Bluetooth are targeting for short-range communications, while LoRaWAN, SIGFOX, Ingenu or Weightless are designed for long-range communications.
We refer to the survey \cite{Centenaro16} for more details, and references in \cite{Azari18} or our recent article \cite{MoyBesson2019}.

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.65\textwidth,angle=90]{United_States_Frequency_Allocations_Chart_2016_The_Radio_Spectrum.pdf}
    \caption[A chart representing the allocation of radio spectrum in the USA in $2016$]{A chart representing the allocation of radio spectrum in the USA in $2016$. Only a small share of the spectrum is not yet allocated, the top-left corner in white (it is in logarithmic scale from \SI{0}{\kilo\hertz} to \SI{300}{\giga\hertz}). \copyright{} Wikimedia.}
    % https://upload.wikimedia.org/wikipedia/commons/c/c7/United_States_Frequency_Allocations_Chart_2016_-_The_Radio_Spectrum.pdf?uselang=fr
    \label{fig:1:United_States_Frequency_Allocations_Chart_2016_The_Radio_Spectrum}
\end{figure}

Cellular network bands are overloaded in most parts of the world, but other frequency bands (such as military, amateur radio and paging frequencies) are insufficiently utilized.
Independent studies performed in some countries confirmed that observation, and concluded that spectrum usage highly depends on time and place.
Moreover, the fixed spectrum allocation prevents rarely used frequencies, like those assigned to specific services, from being used, even when any unlicensed users would not cause noticeable interference to the assigned service.
Therefore, in the last $15$ years, regulatory bodies in the world have been considering whether to allow unlicensed users in licensed bands, if they would not cause any interference to the (paying) licensed users.
These initiatives have initiated the research on \textbf{Cognitive Radio},
and especially on a field that we detail below, \textbf{Dynamic Spectrum Access (DSA)}.

@@

\paragraph{From Statistics or Machine Learning to Multi-Armed Bandits.}
%
We can narrow down the first field of study of this thesis, steps by steps.
% Stats > sequential learning > bandits
A first point-of-view is the historical one, as Multi-Armed Bandits (MAB) learning emerged as an interesting problem in the statistics community and among researchers interested by sequential learning, as shown by the earlier works of \cite{Thompson33,Robbins52,LaiRobbins85}.
% ML > RL > MAB
We can also expose another point-of-view, which starts from the general concept of Machine Learning (ML), and focus on Reinforcement Learning (RL), then on MAB learning.
As illustrated in Figure~\ref{fig:1:MainTypesOfMachineLearning}, there are other types of ML, that we do not focus on in this thesis.
%
The reference book on RL is \cite{SuttonBarto2018}, so let us quote the definition of RL given by R. Sutton and A. Barto:
\emph{``Reinforcement learning is learning what to do - how to map situations to actions - so
as to maximize a numerical reward signal. The learner is not told which actions to
take, but instead must discover which actions yield the most reward by trying them''}.
% - apprentissage statistique (machine learning) > séquentiel > par renforcement > information partielle (bandits)

@@

We illustrate this idea of a \emph{learning cycle} alternating between action and feedback,
in the following Figure~\ref{fig:1:ReinforcementLearningCycle}.
A player (or learner) interacts with its environment by taking an action $A(t)$ (\eg, a choice in a finite set, $A(t)\in\{1,\dots,K\}$, or a vector $A(t)\in\R^d$), and then observing a reward $r(t)$, which is a certain measure of success from the environment (\eg, $r(t)\in\{0,1\}$ for binary failure/success, or $r(t)\in\R$).
The goal of the player is to maximize its reward, by trials and errors (\ie, actions and rewards).
Many real-world problems can be framed as Reinforcement Learning problems, as illustrated by the survey \cite{bouneffouf2019survey} and Section~\ref{sec:2:applicationsofStochasticMAB}, for instance learning to walk or to drive, learning to play a board or computer game, or discovering which treatment is efficient in healing a certain disease (clinical trial), etc.

@@

% Il faut expliquer ce que c'est que les bandits, rapidement, et leur application à l'OSA
As this thesis focus on Reinforcement Learning models,
it is important to highlight that in such decision making models, the learner does not have access to the entire reaction of the environment after taking its action.
%
In other words, the player only sees the reward given by its action at every round, and not the reward that would have been given if she chose any other action.
This kind of limited feedback is called \emph{bandit information}, and we discuss the history and the concept of \emph{Multi-Armed Bandits} (MAB) in details in the next Chapter~\ref{chapter:2}.
Clinical trials and treatment discovery were historically the first applications of MAB since $1930$s, with the early work of W. Thompson \cite{Thompson33}.
%
MAB are a simple yet powerful example of the well-known exploration/exploitation dilemma:
when facing a set of $K$ actions whose effects on the environment are unknown, a learner must balance between
exploration of the unknown actions, in order to collect more information about them,
and exploitation of the best action, according to its current knowledge.
%
The MAB problem have been studied in both the machine learning and the statistics community, since the $1950$s with pioneers like H. Robbins \cite{Robbins52}, and more recently it is an active field of research, since the late $1990$s \cite{Anantharam87a,Anantharam87b,auer1995gambling,Agrawal95}.
Research on MAB has produced a vast literature in the $2000$s \cite{Auer02,Auer02NonStochastic,Audibert2009minimax} and continues to be a topic of high interest, as illustrated by the surveys and books \cite{Bubeck12,LattimoreBanditAlgorithmsBook,Slivkins2019}, and the wide range of applications of MAB in the recent years.

@@

% \subsection{Focusing on IoT networks.}
\subsection{Our models of IoT networks and decentralized MAB}

@@

On the one hand, in the OSA setting, we failed to obtain positive result for the Selfish policy, as we proved that in some limited settings (\eg, $K=3$ and $M=2$), \Selfish-\UCB{} can suffer from linear regret with a small probability, and thus suffers from linear mean regret.
This result was later confirmed and analyzed by \cite{BoursierPerchet18}.
%
On the other hand, we were able to propose new algorithms for this multi-player bandit problem with sensing information, and we analyzed our proposal \MCTopM-\klUCB, to show that it achieves a finite-time logarithmic regret upper bound, improving over previous state-of-the-art results.
Our algorithm also achieves a logarithmic number of collisions and arm switches, and allows a fixed group of $M$ devices to efficiently learn to use the $M$ best channels orthogonally for almost all their up-link communications.
%
This strong theoretical result heavily depends on the presence of sensing feedback, and as such our results are not (yet) applicable to the IoT model without sensing.

@@

On the one hand, if the activation pattern of all devices could be fixed, for instance based on a centralized affectation of the devices to different time slots, then the model of multi-player bandit from Chapter~\ref{chapter:5} can be used to let all groups of $M \leq K$ devices learn an optimal orthogonal affectation in the $K$ channels (\eg, groups can be the set of devices that all emit at the same time).
%
However, even if we continue to assume a synchronized time in all the rest of the thesis,
it is hard to argue that this hypothesis of a centralized time schedule for the end-devices can be realistic, because we study the case of decentralized learning for OSA and IoT precisely in order to avoid any central control of the devices by the base station, as we explained above.

@@

\TODOL{Guy Carrault me dit : Sinon les publications du doctorant sont indiquées en début de document, je ne me rappelle plus si elles ne doivent pas figurées à la fin}

@@

\begin{itemize}
\item
    \emph{Decentralized Spectrum Learning for IoT Wireless Networks Collision Mitigation},
    by Christophe Moy \& \textbf{Lilian Besson}.
    1st International ISIoT workshop,
    % \footnote{~See \href{https://sites.google.com/view/ISIoT2019}{\texttt{sites.google.com/view/ISIoT2019}}},
    at \emph{Conference on Distributed Computing in Sensor Systems},
    % \footnote{~IEEE DCOSS 2019, see \href{http://2019.dcoss.org}{\texttt{2019.dcoss.org}}},
    Santorini, Greece, May $2019$.
    % \href{https://HAL.Inria.fr/hal-02144465}{\texttt{HAL.Inria.fr/hal-02144465}}.
    See Section~\ref{sec:4:gnuradio}.
    \cite{MoyBesson2019}

@@

\item
    \emph{Upper-Confidence Bound for Channel Selection in LPWA Networks with Retransmissions},
    by Rémi Bonnefoi, \textbf{Lilian Besson}, J. Manco-Vasquez \& Christophe Moy.
    1st International MOTIoN workshop,
    % \footnote{~MOTIoN 2019, see \href{https://sites.google.com/view/wcncworkshop-motion2019}{\texttt{sites.google.com/view/wcncworkshop-motion2019}}},
    at \emph{IEEE WCNC}, Marrakech, Morocco, April $2019$.
    % \href{https://HAL.Inria.fr/hal-02049824}{\texttt{HAL.Inria.fr/hal-02049824}}.
    See Section~\ref{sec:4:retransmissions}.
    \cite{Bonnefoi2019WCNC}

\item
    \emph{GNU Radio Implementation of MALIN: ``Multi-Armed bandits Learning for Internet-of-things Networks''},
    by \textbf{Lilian Besson}, Rémi Bonnefoi \& Christophe Moy.
    \emph{Wireless Communication and Networks Conference},
    % \footnote{~IEEE WCNC 2019, see \href{http://wcnc2019.ieee-wcnc.org}{\texttt{wcnc2019.ieee-wcnc.org}}},
    Marrakech, Morocco, April $2019$,
    % \href{https://HAL.Inria.fr/hal-02006825}{\texttt{HAL.Inria.fr/hal-02006825}}.
    See Section~\ref{sec:4:gnuradio}.
    \cite{Besson2019WCNC}

\item
    \emph{Multi-Player Bandits Revisited},
    by \textbf{Lilian Besson} \& Emilie Kaufmann.
    \emph{Algorithmic Learning Theory},
    % \footnote{~ALT 2018, see \href{http://www.cs.cornell.edu/conferences/alt2018}{\texttt{www.cs.cornell.edu/conferences/alt2018}}},
    Lanzarote, Spain, April $2018$,
    % \href{https://HAL.Inria.fr/hal-01629733}{\texttt{HAL.Inria.fr/hal-01629733}}.
    See Chapter~\ref{chapter:5}.
    \cite{Besson2018ALT}

\item
    \emph{Aggregation of Multi-Armed Bandits learning algorithms for Opportunistic Spectrum Access},
    by \textbf{Lilian Besson}, Emilie Kaufmann \& Christophe Moy.
    \emph{Wireless Communication and Networks Conference},
    % \footnote{~IEEE WCNC 2018, see \href{http://wcnc2018.ieee-wcnc.org}{\texttt{wcnc2018.ieee-wcnc.org}}},
    Barcelona, Spain, April $2018$,
    % \href{https://HAL.Inria.fr/hal-01705292}{\texttt{HAL.Inria.fr/hal-01705292}}.
    See Chapter~\ref{chapter:25}.
    \cite{Besson2018WCNC}

@@

\begin{center}
    \textbf{Copyright $2016$-$2019$, \copyright ~Lilian~Besson.}
\end{center}

### 2019-06-26 2-Chapters/Conclusion/conclusion.tex
Lots of things to do, oh my god.

This manuscript was organized in two parts, Part~\ref{part:Introduction} and Part~\ref{part:MABIOT}.
%
The main problem we studied was \emph{``Can we adapt the decision making tools already successfully applied to Cognitive Radio for Opportunistic Spectrum Access to the specific needs of CR for the (future) Internet of Things networks?''}
%
This document detailed the historical, scientific and technical contexts of the problems we considered for our research, and presented the main contributions produced during the last three years.

@@

We started to present the multi-armed bandit model, with a finite number of arms and real-valued rewards.
Our main focus is on one-dimensional exponential families of distributions, and on stochastic and stationary problems.
The notations used in the manuscript were introduced by showcasing an interactive demonstration\footnote{See \href{https://perso.crans.org/besson/phd/MAB\_interactive\_demo/}{\texttt{perso.crans.org/besson/phd/MAB\_interactive\_demo/}}}.

@@

Finally in Section~\ref{sec:4:retransmissions}, we presented an extension of our model of Low-Power Wide-Area (LPWA) networks based on an ALOHA protocol, again under the hypothesis of a protocol slotted in time and frequency, and assuming a perfect synchronisation.
% , in which dynamic IoT devices can again use machine learning algorithms, to improve their Packet Loss Ratio (PLR) when accessing the network.
If the priority is the Quality of Service (QoS), for instance with renewable energy capabilities, this second model is more appropriate.
We presented and evaluated several heuristics that try to learn how to transmit and retransmit in a smarter way, by using the \UCB{} algorithm for channel selection for the first transmission (first-stage), and different proposals based on \UCB{} for the retransmissions upon collisions (second-stage).
The main novelty of this model is to also learn the most efficient way to retransmit packets upon radio collision, by also using a multi-armed bandit algorithm for this second-stage optimization problem.
%
We showed that incorporating learning for the transmission is needed to achieve optimal performance, with significant gain in terms of successful transmission rate in networks with a large number of devices (up-to $30\%$ in the example network).
Our empirical simulations show that each of our proposed heuristic outperforms a naive random access scheme.
Surprisingly, the simulations showed that a simple \UCB{} learning approach, that simply retransmits in the same channel, turns out to perform almost as well as more complicated heuristics, that can change the channel used for retransmissions.

@@

The second chapter of Part~\ref{part:MABIOT} of this manuscript studied Multi-Players Multi-Arm Bandits models.
%
We present three variants of this model,
with different level of feedback being available to the decentralized players, under which we proposed efficient algorithms.
For the two easiest models --with sensing, our theoretical contribution improves the state-of-the-art upper bounds on the regret.
In the absence of sensing, we also provide some motivation for the practical use of the \Selfish{} heuristic, a simple index policy based on hybrid indices that are directly taking  into account the collision information.
\Selfish{} corresponds to the heuristic used in the first model studied in Chapter~\ref{chapter:4}.
%
We also reviewed various variants of this model, and for some interesting variants we discussed the related literature, which has proven to be very active in the last two years.
For some of these variants, we explained why our approach does not work efficiently without modifications, but we detailed and illustrated how to adapt \MCTopM{} to other settings.
For example, it assumes to know the number of players $M$ before-hand, but we illustrated that a previously introduced technique to estimate $M$ can also be applied to our proposal and give satisfactory empirical performances.
% Further works would be required to adapt the theoretical analysis to these various extensions.

>>>>

This manuscript was organized in two parts, Part~\ref{part:Introduction} and Part~\ref{part:MABIOT}.
%
The main problem we studied was \emph{``Can we adapt the decision making tools already successfully applied to Cognitive Radio for Opportunistic Spectrum Access to the specific needs of CR for the (future) Internet of Things networks?''}
%
We detailed the historical, scientific and technical contexts of the problems we considered for our research, and presented the main contributions produced during the last three years.

@@

We started to present the multi-armed bandit model, with a finite number of arms and real-valued rewards.
Our main focus is on one-dimensional exponential families of distributions, and on stochastic and stationary problems.
The notations used in the manuscript were introduced by showcasing an interactive demonstration.
% \footnote{See \href{https://perso.crans.org/besson/phd/MAB\_interactive\_demo/}{\texttt{perso.crans.org/besson/phd/MAB\_interactive\_demo/}}}.

@@

Finally in Section~\ref{sec:4:retransmissions}, we presented an extension of our model of Low-Power Wide-Area (LPWA) networks based on an ALOHA protocol, again under the hypothesis of a protocol slotted in time and frequency, and assuming a perfect synchronisation.
% , in which dynamic IoT devices can again use machine learning algorithms, to improve their Packet Loss Ratio (PLR) when accessing the network.
If the priority is the Quality of Service (QoS), for instance with renewable energy capabilities, this second model is more appropriate.
We presented and evaluated heuristics that learn how to transmit and retransmit in a smarter way, by using the \UCB{} algorithm for channel selection for the first transmission (first-stage), and different proposals based on \UCB{} for the retransmissions upon collisions (second-stage).
The main novelty of this model is to also learn the most efficient way to retransmit packets upon radio collision, by also using a MAB algorithm for this second-stage optimization problem.
%
We showed that incorporating learning for the transmission is needed to achieve optimal performance, with significant gain in terms of successful transmission rate in networks with a large number of devices (up-to $30\%$ in the example network).
Our empirical simulations show that each of our proposed heuristic outperforms a naive random access scheme.
Surprisingly, the simulations showed that a simple \UCB{} learning approach, that simply retransmits in the same channel, turns out to perform almost as well as more complicated heuristics, that can change the channel used for retransmissions.

@@

The second chapter of Part~\ref{part:MABIOT} of this manuscript studied Multi-Players Multi-Arm Bandits models.
%
We present three variants of this model,
with different level of feedback being available to the decentralized players, under which we proposed efficient algorithms.
For the two easiest models --with sensing, our theoretical contribution improves the state-of-the-art upper bounds on the regret.
In the absence of sensing, we also provide some motivation for the practical use of the \Selfish{} heuristic, a simple index policy based on hybrid indices that are directly taking  into account the collision information.
\Selfish{} corresponds to the heuristic used in the first model studied in Chapter~\ref{chapter:4}.
%
We also reviewed various variants of this model, and for some interesting variants we discussed the related literature, which has proven to be very active in the last two years.
For some of these variants, we explained why our approach does not work efficiently without modifications, but we detailed and illustrated how to adapt \MCTopM{} to other settings.
For example, it assumes to know the number of players $M$ before-hand, but we illustrated that a previously introduced technique to estimate $M$ can also be applied successfully to our proposal.
%  and give satisfactory empirical performances.
% Further works would be required to adapt the theoretical analysis to these various extensions.
